Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-11-06 paper/paper.tex
Rewording a bit of Section 2

\commentsfalse        %% toggle comments here
\extendedfalse
\anonymoustrue

@@

\ifcomments
\newnote{scw}{blue} % Stephanie Weirich
\newnote{yl}{purple} % Yiyun Liu
\newnote{jc}{darkgreen}
\else
\newcommand{\scw}[1]{}
\newcommand{\yl}[1]{}
\newcommand{\jc}[1]{}
\fi

@@

\begin{figure}[H]
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
\caption{Parallel reduction ($\beta$ rules only) }
\label{fig:par}
\end{figure}

Before we specify the typing rules, we first specify the equational theory
used in the conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). The
equivalence relation used in this rule is often referred to
\textit{definitional equality} in dependent type theories because it defines
the equivalence that the syntactic type system works up to.

In \lang{}, we use a relation called \emph{convertibility} for definitional
equality. Two terms are convertible, if they reduce to a common form. The
reduction that we use is called \emph{parallel reduction}, written
$[[a => b]]$. The notation $[[a =>+ b]]$ indicates the transitive closure of
parallel reduction.

@@

The definition of the parallel reduction relation, appears in
Figure~\ref{fig:par}. (For brevity, the reflexivity and congruence rules of
this relation are omitted from this figure).

We prove, through standard
techniques~\citet{takahashi-parallel-reduction,plfa22.08}, the following
properties of parallel reduction.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 / x } => a1 {
    b1 / x }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b / x } => a1 {b / x}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}

@@

\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}

The convertibility relation that we use for conversion in \lang{} is unusual
in that it is directly defined via parallel reduction, instead of using the
related notion of $\beta$-equivalence~\citep{barendregt1991introduction,coquand1990:cic}. This choice
does not change the language definition; a detailed argument of the
equivalence between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found
in \citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}. However, this choice simplifies later
proofs, as we discuss in Section~\ref{sec:discuss}.

Our definition of equality is untyped: the judgement does not require the two
terms to type check and have the same type. The use of an untyped relation for
conversion is similar to Barendregt's Pure Type
Systems~\cite{barendregt1991introduction} and differs
from MLTT~\citep{Martin-Lof-1973}, where the judgmental
equality takes the form $\Gamma \vdash a \equiv b : A$.
% from which one can
% usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
% reduction.
By working with an untyped judgement, we can establish its
properties independently from the type system and the logical relation, using
well-established syntactic approaches.
%
\citet{siles2012pure} show the equivalence of Barendregt's Pure Type
System, which employs untyped equality, and
its variant that uses typed judgmental equality. This assures us that we do
not lose generality working with a system with untyped
conversion.
\yl{removed the eta equivalence sentence because it might give the
  wrong message that we extended Siles' work with eta:
 Furthermore, we discuss how this definition can be extended with
 $\eta$-equivalence of functions in Section~\ref{XXX} }
We compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

@@

\begin{figure}[H]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{}
\[ \drule{Ctx-Empty} \qquad \drule[width=3in]{Ctx-Cons} \]
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
 Void, Bool, True, False}
\[ \drule[width=5in]{T-If}\]
\[ \drule[width=3in]{T-Eq} \qquad \drule{T-Refl} \]
\[ \drule[width=4in]{T-J} \]
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

Figure~\ref{fig:typing} gives the full typing rules for \lang{}. The premises
wrapped in \colorbox{lightgray}{gray} boxes can be shown to be admissible
syntactically, though some of them are required to strengthen the inductive
hypothesis of the fundamental theorem.

The typing rules of \lang{} are standard for dependent
type theories.  The variable rule, \rref{T-Var}, uses the auxiliary relation
$[[x : A in G]]$, that holds when a variable declaration is found in the
typing context.  The typing of universes ensures that each one belongs to the
next higher level.  \Rref{T-Pi} ensures predicative quantification by
requiring that all parts of the type be typeable at the same universe
level. \Rref{T-Abs} ensures that all functions have well-formed dependent
types. In an application (\rref{T-App}) the argument is substituted for the
variable in the result type.

\Rref{T-Conv} uses the convertibility relation from earlier
as our equality judgment for type conversion.

\scw{Cassia did not follow our discussion of dependent pattern matching. We need to expand}
The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
matching.  The result type of this expression, $[[A {a / x}]]$, is composed of some
\emph{motive} $[[A]]$, a type where its single free variable has been replaced
with the condition of the if expression. When typing the true branch, this
substitution replaces the variable by $[[true]]$, and similarly for the false
branch. As a result, the type system communicates the information gained from
the test to each of the branches of the expression.

A similar sort of dependent pattern matching occurs when eliminating identity
types. Such types are checked for well-formedness with \rref{T-Eq} and
introduced by \rref{T-Refl}. In \rref{T-J}, the elimination form, the subterm
$p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
is the body of the elimination form. In this rule, $[[B]]$ is the motive and
has two free variables.  When checking $[[c]]$, the substitution for these
variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
witnessing the information gained through dependent pattern matching.

The universe hierarchy and the boolean base type gives \lang{} the
ability to compute a type using a term as input, a feature commonly referred to
as large elimination. For
example, we can type check the function $[[\ x . if x Bool Void]]$,
which returns either $[[Bool]]$ or $[[Void]]$ depending on whether its
input is $[[true]]$ or $[[false]]$.
% Working with a system with untyped equality has the huge benefit that
% the confluence result for untyped parallel reduction
% (Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
% resort to the complex syntatic (resp. semantic) technique from
% \citet{siles2012pure} (resp. \citet{decagda}) to resolve
% the circularity of subject reduction and $\Pi$-injectivity.
% Section~\ref{sec:extension} explains how we
% generalize our technique to include $\eta$-law for functions and
% show the existence of normal form for well-typed (open and closed)
% terms, achieving a similar level of expressiveness of the type system
% and strength of metatheoretic property as \citet{decagda}.

>>>>

\commentstrue         %% toggle comments here
\extendedfalse
\anonymousfalse

@@

\ifcomments
\newnote{scw}{blue} % Stephanie Weirich
\newnote{yl}{purple} % Yiyun Liu
\newnote{jc}{red} % Jonathan Chan
\else
\newcommand{\scw}[1]{}
\newcommand{\yl}[1]{}
\newcommand{\jc}[1]{}
\fi

@@

Before we specify the typing rules, we first specify the equational theory
used in the conversion \rref{T-Conv} in Figure~\ref{fig:typing},
known as \textit{definitional equality} in dependent type theories because it
defines the equivalence that the syntactic type system works up to. 
The definitional equality we use is \emph{convertibility}:
two terms are convertible if they reduce to a common form.
The reduction that we use is \emph{parallel reduction}, written
$[[a => b]]$, with $[[a =>+ b]]$ indicating its reflexive, transitive closure.
The parallel reduction relation is defined in Figure~\ref{fig:par},
which omits reflexivity and congruence rules for brevity.

@@

\begin{figure}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
\caption{Parallel reduction ($\beta$ rules only) }
\label{fig:par}
\end{figure}

We prove, through standard
techniques~\citep{takahashi-parallel-reduction,plfa22.08}, the following
properties of parallel reduction.

\begin{lemma}[Reflexivity (parallel reduction)\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Congruence (p.r.)\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 / x } => a1 {
    b1 / x }]]$.
\end{lemma}
\begin{corollary}[Substitution (p.r.)\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b / x } => a1 {b / x}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Diamond property\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}

@@

\begin{lemma}[Reflexivity (convertibility)\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Symmetry (convertibility)\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Transitivity (convertibility)\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}

The convertibility relation that we use for definitional equality is unusual
in that it is directly defined via parallel reduction, instead of using the
related notion of $\beta$-equivalence~\citep{barendregt1991introduction,coquand1990:cic}.
This choice does not change the language definition; a detailed argument of the
equivalence between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found
in \citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}. However, this choice simplifies later
proofs, as we discuss in Section~\ref{sec:discuss}.

Our definitional equality is untyped: the judgment does not require the two
terms to type check and have the same type. The use of an untyped relation for
conversion is similar to Barendregt's Pure Type
Systems~\cite{barendregt1991introduction} and differs
from MLTT~\citep{Martin-Lof-1973}, where definitional
equality takes the form $\Gamma \vdash a \equiv b : A$.
% from which one can
% usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
% reduction.
By working with an untyped judgment, we can establish its
properties independently of the type system and the logical relation, using
well-established syntactic approaches.
%
\citet{siles2012pure} show the equivalence of Barendregt's Pure Type
System, which employs untyped equality, and
its variant that uses typed equality. This assures us that we do
not lose generality working with a system with untyped
conversion.
\jc{I'm not too fond of this last sentence because we \emph{do} lose
the ability to add typed $\eta$ equivalences.}
\yl{removed the eta equivalence sentence because it might give the
  wrong message that we extended Siles' work with eta:
 Furthermore, we discuss how this definition can be extended with
 $\eta$-equivalence of functions in Section~\ref{XXX} }
We compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

@@

\begin{figure}
\begin{mathpar}
  \fbox{$[[ |- G]]$} \hfill \textit{(Context well-formedness)} \\
  \drule{Ctx-Empty} \drule[width=3in]{Ctx-Cons}
\end{mathpar}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Void, Absurd, Bool, True, False}
\begin{mathpar}
  \drule[width=3in]{T-If} \drule[width=2in]{T-Eq} \\
  \drule{T-Refl} \drule[width=4in]{T-J}
\end{mathpar}
\caption{Syntactic typing}
\label{fig:typing}
\end{figure}

Figure~\ref{fig:typing} gives the complete typing rules. The premises
highlighted in \colorbox{lightgray}{gray} can be shown to be admissible
syntactically, though they are required to strengthen the inductive
hypothesis of the fundamental theorem.

These rules are standard for dependent
type theories.  The variable rule, \rref{T-Var}, uses the auxiliary relation
$[[x : A in G]]$ that holds when a variable declaration is found in the
typing context. \Rref{T-Set} ensures that each universe belongs to the
next higher level. \Rref{T-Pi} ensures predicative quantification by
requiring that the domain and codomain types be typeable at the same universe
level. \Rref{T-Abs} ensures that all functions have well-formed dependent
types. In the application \rref{T-App}, the argument is substituted for the
variable in the result type.
\Rref{T-Conv} uses the convertibility relation from earlier
as our equality judgment for type conversion.

\scw{Cassia did not follow our discussion of dependent pattern matching. We need to expand}
The elimination form for booleans in \rref{T-If} demonstrates dependent pattern
matching: the type of the expression \emph{depends} on the term being matched.
This result type involves a \emph{motive} $[[A]]$ abstracted over a boolean.
For the overall expression, $[[x]]$ is filled in by the match target $[[a]]$,
while in the $[[true]]$ (resp. $[[false]]$) branch, $[[x]]$ is filled in by
$[[true]]$ (resp. $[[false]]$). Informally, the type of each branch is more
precise than those in a nondependent match expression, since it knows which
branch it is in. An alternative perspective is that to prove $[[A {a/x}]]$, it
suffices to show that it holds for the two canonical cases that $[[a]]$ is
$[[true]]$ or is $[[false]]$.

% The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
% matching.  The result type of this expression, $[[A {a / x}]]$, is composed of some
% \emph{motive} $[[A]]$, a type where its single free variable has been replaced
% with the condition of the if expression. When typing the true branch, this
% substitution replaces the variable by $[[true]]$, and similarly for the false
% branch. As a result, the type system communicates the information gained from
% the test to each of the branches of the expression.

Dependent pattern matching similarly occurs when eliminating equality types.
Their well-formedness in \rref{T-Eq} enforces that the endpoints of the
equality have the same type, and there is a single canonical proof introduced
by \rref{T-Refl}. The elimination form in \rref{T-J} takes a proof of an
equality $[[p]]$ between $[[a]]$ and $[[b]]$, as well as an elimination body
$[[c]]$. Here, the motive $[[B]]$ is abstracted over the right endpoint $[[x]]$
and an equality $[[y]]$ between it and the fixed left endpoint $[[a]]$. The
eliminator states that to prove $[[B{b, p / x, y}]]$, it suffices to prove the
canonical case that $[[p]]$ is $[[refl]]$ (and consequently that $[[b]]$ is
$[[a]]$), which is witnessed by $[[c]]$.

% In \rref{T-J}, the elimination form, the subterm
% $p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
% is the body of the elimination form. In this rule, $[[B]]$ is the motive and
% has two free variables.  When checking $[[c]]$, the substitution for these
% variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
% witnessing the information gained through dependent pattern matching.

The universe hierarchy and the boolean base type give the
ability to compute a type using a term as input, referred to as
\emph{large elimination}. For example, the well-typed function
$[[\ x . if x Bool Void]]$ returns either $[[Bool]]$ or $[[Void]]$
depending on whether its input is $[[true]]$ or $[[false]]$.
% Working with a system with untyped equality has the huge benefit that
% the confluence result for untyped parallel reduction
% (Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
% resort to the complex syntatic (resp. semantic) technique from
% \citet{siles2012pure} (resp. \citet{decagda}) to resolve
% the circularity of subject reduction and $\Pi$-injectivity.
% Section~\ref{sec:extension} explains how we
% generalize our technique to include $\eta$-law for functions and
% show the existence of normal form for well-typed (open and closed)
% terms, achieving a similar level of expressiveness of the type system
% and strength of metatheoretic property as \citet{decagda}.
